Examining FDR correction issue in clusterProfiler - RNA-seq example

### 2024-09-06 analysis/analyse_rnaseq1.Rmd
add whiskers

colMeans(itxlenmx,na.rm=TRUE) / colMeans(speclenmx,na.rm=TRUE)

mean( ( colMeans(speclenmx,na.rm=TRUE) - colMeans(itxlenmx,na.rm=TRUE) ) / colMeans(itxlenmx,na.rm=TRUE) )

@@

>>>>

rowMeans(itxlenmx,na.rm=TRUE)

rowMeans(speclenmx,na.rm=TRUE)

( rowMeans(speclenmx,na.rm=TRUE) - rowMeans(itxlenmx,na.rm=TRUE) ) / rowMeans(itxlenmx,na.rm=TRUE)

mean( ( rowMeans(speclenmx,na.rm=TRUE) - rowMeans(itxlenmx,na.rm=TRUE) ) / rowMeans(itxlenmx,na.rm=TRUE) )

@@

rowMeans( ( res2b - res2a ) / res2a ) * 100

### 2024-09-06 analysis/analyse_rnaseq2.Rmd
add whiskers

@@

@@

@@

library(reshape2)
res2a <- as.matrix(acast(res1, Var1~Var2, fun.aggregate = sum, value.var="nsig_bad"))
rownames(res2a) <- gsub("bulkrna","d",rownames(res2a))
rownames(res2a) <- gsub(".Rds"," original",rownames(res2a))

res2b <- as.matrix(acast(res1, Var1~Var2, fun.aggregate = sum, value.var="nsig_good"))
rownames(res2b) <- gsub("bulkrna","d",rownames(res2b))
rownames(res2b) <- gsub(".Rds"," corrected",rownames(res2b))

res2 <- rbind(res2a,res2b)
res2 <- res2[order(rownames(res2)),]
res2c <- rowMeans(res2)
res2 <- cbind(res2,res2c)

res2 %>%
  kbl(row.names = TRUE, caption="Effect of FDR problem correction across seven studies and nine gene set libraries") %>%
  kable_paper("hover", full_width = F)

saveRDS(res2,"fdrprob.rds")

>>>>

jac3 <- do.call(rbind,lapply(jac,function(x) {x[,"GO"] } ))

apply(jac3,1,function(x) { mean(x, na.rm=TRUE) } )
apply(jac3,1,function(x) { sd(x, na.rm=TRUE) } )

jsd <- lapply(jac,function(y) {
  res <- apply(y,2,function(x) {
    x <- x[!is.na(x)]
    x <- x[is.finite(x)]
    sd(x)
  } )
  return(res)
})

jsd <- do.call(rbind,jsd)

jsd %>%
  kbl(row.names = TRUE, caption="Standard deviation in Jaccard stats across gene list length") %>%
  kable_paper("hover", full_width = F)

@@

jac %>%
  kbl(row.names = TRUE, caption="Jaccard stats across gene list length") %>%
  kable_paper("hover", full_width = F)

# GO with error bars
plot(jac[,"GO"],type="b", pch=19,cex=1.5,lwd=2,
  xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected")
axis(1, at = seq(1, 5, by = 1),
     labels = ngenes)
mysd <- jsd[,"GO"]
arrows(x0=1:5, y0=(jac[,"GO"] - mysd) ,  x1=1:5, y1=(jac[,"GO"] + mysd), code=3, angle=90, length=0.1 )

mysd <- jsd[,"Reactome"]
points(jac[,"Reactome"],type="b",col="red",pch=19,cex=1.5,lwd=2)
arrows(x0=1:5, y0=(jac[,"Reactome"] - mysd) ,  x1=1:5, y1=(jac[,"Reactome"] + mysd), code=3, angle=90, length=0.1 ,col="red")

mysd <- jsd[,"Cellmarkers"]
points(jac[,"Cellmarkers"],type="b",col="blue",pch=19,cex=1.5,lwd=2)
arrows(x0=1:5, y0=(jac[,"Cellmarkers"] - mysd) ,  x1=1:5, y1=(jac[,"Cellmarkers"] + mysd), code=3, angle=90, length=0.1 , col="blue")

mysd <- jsd[,"TFT GTRD"]
points(jac[,"TFT GTRD"],type="b",col="gray50",pch=19,cex=1.5,lwd=2)
arrows(x0=1:5, y0=(jac[,"TFT GTRD"] - mysd) ,  x1=1:5, y1=(jac[,"TFT GTRD"] + mysd), code=3, angle=90, length=0.1 , col="gray50" )

@@

Table when using 2000 genes.

@@

library(reshape2)
res2000 <- subset(res1,Var3==2000)

res3a <- as.matrix(acast(res2000, Var1~Var2, fun.aggregate = sum, value.var="nsig_bad"))
rownames(res3a) <- gsub("bulkrna","d",rownames(res3a))
rownames(res3a) <- gsub(".Rds"," original",rownames(res3a))

res3b <- as.matrix(acast(res2000, Var1~Var2, fun.aggregate = sum, value.var="nsig_good"))
rownames(res3b) <- gsub("bulkrna","d",rownames(res3b))
rownames(res3b) <- gsub(".Rds"," corrected",rownames(res3b))

res3 <- rbind(res3a,res3b)
res3 <- res3[order(rownames(res3)),]
res3c <- rowMeans(res3)
res3 <- cbind(res3,res3c)

res3 %>%
  kbl(row.names = TRUE, caption="Effect of FDR problem correction across seven studies and nine gene set libraries") %>%
  kable_paper("hover", full_width = F)

saveRDS(res3,"fdrprob.rds")

Another table showing downsampling result.

```{r,dstable}

ds <- res1

colnames(ds) <- c("Dataset","GeneSetLibrary","GeneListSize","nSig_original","nSig_corrected","Jaccard")

ds %>%
  kbl(row.names = TRUE, caption="FDR problem worsens when the input gene list is shorter") %>%
  kable_paper("hover", full_width = F)

saveRDS(ds,"downsample_fdr.rds")
